---
title: "Expected Revenue from NBA Playoffs"
author: "Dan Salo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `files` code block reads in the wins and revenue data.

```{r files}
revs <- read.csv("/abs/path/to/NBA-Analytics/gate-revenue/data/revs.csv", header=TRUE, sep=",")
wins <- read.csv("/abs/path/to/NBA-Analytics/gate-revenue/data/wins.csv", header=TRUE, sep=",")
```

The `gate-revenue-plot` code block plots a grouped bar chart dislaying the expected gate revenue at NBA stadiums for each round of the playoffs.

```{r gate-revenue-plot, message=FALSE, warning=FALSE}
library(plotly)
plot_ly(revs, x = ~HomeTeam, y = ~Round1_Gate, type = 'bar', name = 'Round 1', marker = list(color = 'rgb(0, 12195182)')) %>%
  add_trace(y = ~Round2_Gate, name = 'Round 2', marker = list(color = 'rgb(240, 81, 51)')) %>%
  add_trace(y = ~Round3_Gate, name = 'Round 3', marker = list(color = 'rgb(253, 187, 48)')) %>%
  add_trace(y = ~Round4_Gate, name = 'Round 4', marker = list(color = 'rgb(0, 45, 98)')) %>%
  layout(title = 'Expected Gate Revenue at NBA Stadiums',
         xaxis = list(
           title = "",
           tickfont = list(
             size = 14,
             color = 'rgb(107, 107, 107)')),
         yaxis = list(
           title = 'Revenue ($)',
           titlefont = list(
             size = 16,
             color = 'rgb(107, 107, 107)'),
           tickfont = list(
             size = 14,
             color = 'rgb(107, 107, 107)')),
         legend = list(x = 0.8, y = 1, bgcolor = 'rgba(255, 255, 255, 0)', bordercolor = 'rgba(255, 255, 255, 0)'),
         barmode = 'group', bargap = 0.15)
```

The `functions` code block reads in the included CSV files and defines the functions necessary to simulate entire playoff series and determine the amount of revenue produced.

```{r functions}
# We define a function `predict_revenue` that will give the expected revenue and the probability of the home team advancing:
predict_revenue <- function(p_a, p_h, rev) {

  # First, predict the chance of games going to 4,5,6 or 7 games.
  ## Probability of a 7-game series
  p7 = choose(3,2)* choose(3,1) * ( p_h^2*p_a^2*(1-p_h)^2*(1-p_a) + (1-p_a)^2*(1-p_h)^2*p_h^2*p_a) + choose(3,1) * choose(3,2)* ( p_h^3*p_a*(1-p_a)^2*(1-p_h) + (1-p_a)*(1-p_h)^3*p_a^2*p_h ) + ( p_h*p_a^3*(1-p_h)^3 + (1-p_a)^3*(1-p_h)*p_h^3 ) + ( p_h^4*(1-p_a)^3 + (1-p_h)^4*p_a^3 )

  ## Probability of a 6-game series
  p6 = choose(2,1)* choose(3,1)* ( p_h^2*p_a^2*(1-p_h)*(1-p_a) + (1-p_a)^2*(1-p_h)^2*p_h*p_a) + ( p_h^3*p_a*(1-p_a)^2 + (1-p_a)*(1-p_h)^3*p_a^2 ) + choose(3,2) * ( p_h*p_a^3*(1-p_h)^2 + (1-p_a)^3*(1-p_h)*p_h^2 )

  ## Probability of a 5-game series
  p5 = choose(2,1) * (p_h^2*p_a^2*(1-p_h) + (1-p_a)^2*(1-p_h)^2*p_h) + choose(2,1) * ( p_h^3*p_a*(1-p_a) + (1-p_a)*(1-p_h)^3*p_a)

  ## Probability of a 4-game series
  p4 = (p_h^2*p_a^2 + (1-p_a)^2*(1-p_h)^2)

  ## Expected value computation
  money = p4 * 4 * rev + p5 * 5* rev + p6 * 6 * rev + p7 * 7 * rev

  # Next, predict the winner
  home <- 0
  ## Probability home advancing after 7
  home <- home + choose(3,2)* choose(3,1) * p_h^2*p_a^2*(1-p_h)^2*(1-p_a) +  choose(3,1) * choose(3,2)* p_h^3*p_a*(1-p_a)^2*(1-p_h) + p_h*p_a^3*(1-p_h)^3 + p_h^4*(1-p_a)^3

  ## Probability home advancing after 6
  home <- home + choose(2,1)* choose(3,1)* p_h^2*p_a^2*(1-p_h)*(1-p_a) + p_h^3*p_a*(1-p_a)^2 + choose(3,2) * p_h*p_a^3*(1-p_h)^2

  ## Probability home advancing after 5
  home <- home + choose(2,1) * p_h^2*p_a^2*(1-p_h) + choose(2,1) * p_h^3*p_a*(1-p_a)

  ## Probability home advancing after 4
  home <- home + p_h^2*p_a^2

  return(c(money, home))
}

#We need to define a function that determine which team is home and away, assuming the if two teams with the same seed from different conferences meet then the team from the West has the home court advantage:

home_away <- function(team1, team2) {
  # Preset the home/away teams
  home <- team1
  away <- team2

  # Extract seed and conference from string
  num1 <- as.numeric(substr(team1,nchar(team1), nchar(team1)))
  conf1 <- substr(team1,1, 4)
  num2 <- as.numeric(substr(team1,nchar(team1), nchar(team1)))
  conf2 <- substr(team2,1, 4)

  # Do logic to send back home/away team
  if (conf1 != conf2) {
    if (num1 == num2) {
      if (conf1 == 'East') {
        home <- team2
        away <- team1
      }
    }
  }
  if (num1 < num2) {
    home <- team1
    away <- team2
  }

  return(list('home' = home, 'away' = away))
}

#Next, we wrap `predict_revenue` in another function called `simulate_round` that will analyze all the matchups within that round for one conference:

simulate_conf_round <- function(round, teams) {
  round_rev <- 0
  winners <- c()
  for (e in seq(nrow(teams))) {
    # Get wins data
    team1 <- teams[e,1]
    team2 <- teams[e,2]
    sorted_teams <- home_away(team1, team2)
    row <- wins[ which(wins$Team1 == sorted_teams$home & wins$Team2 == sorted_teams$away), ]
    p_h <- row$Prob1WinsHome
    p_a <- row$Prob1WinsAway

    # Get gate revenue datas
    rev <- as.numeric(revs[ which(revs$HomeTeam == sorted_teams$home),][sprintf('Round%d_Gate', round)])

    # Update revenue for round
    result <- predict_revenue(p_h, p_a, rev)
    round_rev <- round_rev + result[1]

    # Update winners circle
    winner <- NULL
    if (runif(1,0,1) < result[2]) {
      winner <- sorted_teams$home
    } else {
      winner <- sorted_teams$away
    }
    winners <- c(winners, winner)
  }
  return(list('rev' = round_rev, 'winners' = winners))
}

# Since rounds 1,2,3 only involve teams within the conference, we can automate those simulations using a general function.
simulate_full_round <- function(round, state, print_bool=FALSE) {

  ## Simulate both conferences
  resultEast = simulate_conf_round(round, state$east_teams[[round]])
  resultWest = simulate_conf_round(round, state$west_teams[[round]])

  ## Update State
  row = c(2,1,1)
  state$east_teams[[round+1]] <- matrix(resultEast$winners, nrow=row[round], byrow=TRUE)
  state$west_teams[[round+1]] <- matrix(resultWest$winners, nrow=row[round], byrow=TRUE)
  state$west_revs[round] <- resultWest$rev
  state$east_revs[round] <- resultEast$rev

  ## Print Totals
  if (print_bool) {
    sprintf('Western Conference Round 1 Revenue (mil): %f', resultWest$rev / 10^6)
    sprintf('Eastern Conference Round 1 Revenue (mil): %f', resultEast$rev / 10^6)
    sprintf('Total Round 1 Revenue (mil): %f', (resultWest$rev + resultEast$rev) / 10^6)
    sprintf('Total Revenue after Round 1 (mil): %f', sum(RevTotal) / 10^6)
  }

  return(state)
}

simulate_finals <- function(state) {
  playoff_teams = matrix(c(state$east_teams[[4]], state$west_teams[[4]]), nrow=1, byrow=TRUE)
  result = simulate_conf_round(4, playoff_teams)

  ## Update State
  state$finals_rev = result$rev

  return(state)
}

simulate_playoffs <- function() {
  # Define 1st round games and initialize 2nd and 3rd round for East/West
  east1 = matrix(c('East1', 'East8', 'East4', 'East5', 'East3', 'East6', 'East2', 'East7'), ncol=2, byrow=TRUE)
  west1 = matrix(c('West1', 'West8', 'West4', 'West5', 'West3', 'West6', 'West2', 'West7'), ncol=2, byrow=TRUE)

  # Initialize State
  state <- list('east_revs' = c(), 'west_revs' = c(), 'east_teams' = list(east1), 'west_teams' = list(west1), 'finals_revs' = 0)

  # Simulate first 3 rounds
  state = simulate_full_round(1, state)
  state = simulate_full_round(2, state)
  state = simulate_full_round(3, state)

  # Simulate Finals game
  state = simulate_finals(state)
  return(state)
}
```

The `main` code block will simulate $S$ number of playoffs and track the teams that advanced and the money generated.

```{r main}
# Initialize
total <- c()
counts <- list(big = 0, small = 0)
all_teams <- list('East8' = 0, 'East7' = 0, 'East6' = 0, 'East5' = 0, 'East4' = 0, 'East3' = 0, 'East2' = 0, 'East1' = 0, 'West1' = 0, 'West2' = 0, 'West3' = 0, 'West4' = 0, 'West5' = 0, 'West6' = 0, 'West7' = 0, 'West8' = 0)
teams <- list(big = list(round2 = all_teams, round3 = all_teams, round4 = all_teams), small = list(round2 = all_teams, round3 = all_teams, round4 = all_teams))

# Helper function
matrix_expand <- function(mat) {
  output <- c()
  for (r in seq(nrow(mat))) {
    for (c in seq(ncol(mat))) {
      output <- c(output, mat[r,c])
    }
  }
  return(output)
}

team_tally <- function(ste, tms, size) {
  for (round in c(2,3,4)) {
    for (t in matrix_expand(ste$west_teams[[round]])) {
      tms[[size]][[sprintf('round%d', round)]][[t]] <- tms[[size]][[sprintf('round%d', round)]][[t]] + 1
    }
    for (t in matrix_expand(ste$east_teams[[round]])) {
      tms[[size]][[sprintf('round%d', round)]][[t]] <- tms[[size]][[sprintf('round%d', round)]][[t]] + 1
    }
  }
  return(tms)
}

S <- 1000
for (i in seq(S)) {
  state = simulate_playoffs()
  rev = sum(state$east_revs) + sum(state$west_Revs) + state$finals_rev
  total <- c(total, rev)
  if ((115 * 10^6) < rev & rev < (120 * 10^6)) {
    counts$small = counts$small + 1
    teams <- team_tally(state, teams, 'small')
  }
  if ((155 * 10^6) < rev & rev < (165 * 10^6)) {
    counts$big = counts$big + 1
    teams <- team_tally(state, teams, 'big')
  }
}
```

The `expected-revenue-plot` code block will generate a normalized histogram of the possible revenue outcomes after running the simulations.

```{r expected-revenue-plot}
plot_ly(x = total, type = "histogram", histnorm = "probability",
             marker = list(color = 'rgb(0,131,72)',
                           line = list(color = 'rgb(0,0,0)', width = 1.5))) %>%
  layout(title = 'Probability Distribution of Revenue',
           xaxis = list(
             title = "Revenue ($)",
             tickfont = list(
               size = 14,
               color = 'rgb(107, 107, 107)')),
           yaxis = list(
             title = 'Probability',
             titlefont = list(
               size = 16,
               color = 'rgb(107, 107, 107)'),
             tickfont = list(
               size = 14,
               color = 'rgb(107, 107, 107)')))
```

The probability distribution of revenue is bimodal, with the two modes account for about 75\% of the possible outcomes, so we will investigate the teams that advanced in each scenario. The `team-plots` code block will generate a grouped bar chart for each round with the teams that advanced when the total playoff revenue was $160 million and $120 million.
```{r teams-plots}

plot_bar <- function(tms, r) {
  data <- data.frame(t(rbind(data.frame(tms$big[[sprintf('round%d', r)]])/counts$big, data.frame(tms$small[[sprintf('round%d', r)]])/counts$small)))
  names(data) <- c('big','small')
  p <- plot_ly(data, x = names(all_teams), y = ~big, type = 'bar', name = '$160 mil', marker = list(color = 'rgb(0, 107, 182)')) %>%
    add_trace(y = ~small, name = '$120 mil', marker = list(color = 'rgb(134, 0, 56)')) %>%
    layout(title = sprintf('Round %d Teams', r),
           xaxis = list(
             title = "",
             tickfont = list(
               size = 14,
               color = 'rgb(107, 107, 107)')),
           yaxis = list(
             title = 'Percentage of Category',
             titlefont = list(
               size = 16,
               color = 'rgb(107, 107, 107)'),
             tickfont = list(
               size = 14,
               color = 'rgb(107, 107, 107)')),
           legend = list(x = 0.8, y = 1, bgcolor = 'rgba(255, 255, 255, 0)', bordercolor = 'rgba(255, 255, 255, 0)'),
           barmode = 'group', bargap = 0.15)
  return(p)
}
plot_bar(teams, 2)
plot_bar(teams, 3)
plot_bar(teams, 4)
```

We can see from the Round 4 Teams plot that the top team in the Western Conference, the Warriors, need to make it to the Finals if the NBA wants to make around $160 million as opposed to around $120 million.
