---
title: "MC Simulation for Consecutive Losses"
author: "Dan Salo"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The required inputs are `p_win`, which is held at 80\%, `n_games`, which

```{r mc-simulation}
p_win <- 0.8 # probability to win a game
n_games <- 82 # number of games in the season
S <- 100000 # number of monte carlo samples to draw.
last_game <- TRUE
tally <- c()
total_mean <- c()

for (i in seq(0, S-1)) {
  wins = runif(n_games, 0, 1) < p_win
  for (w in wins) {
    if (w == FALSE && last_game == FALSE) {
      tally[i] <- 0
      break
    } else {
      last_game <- w
      tally[i] <- 1
    }
  }
  total_mean <- c(total_mean, sum(tally)/i)
}
```

We can visualize the simulation's progress with a simple line chart:

```{r simulation-plot, message=FALSE, warning=FALSE}
library(plotly)
data <- data.frame(seq(length(total_mean)), total_mean)
names(data) <- c('x','y')
p <- plot_ly(data, x = ~x, y = ~y, name = 'trace 0', type = 'scatter', mode = 'lines', line=list(color='rgb(0,140,168)')) %>%
      layout(title = sprintf('%d Monte Carlo Simulations', S),
         xaxis = list(
           title = "Cumulative Samples",
           tickfont = list(
             size = 14,
             color = 'rgb(107, 107, 107)')),
         yaxis = list(
           title = 'Probability of Losing Consecutive Games',
           titlefont = list(
             size = 16,
             color = 'rgb(107, 107, 107)'),
           tickfont = list(
             size = 14,
             color = 'rgb(107, 107, 107)')))
p
```

By comupting `mean(tally)`, the code predicts that the team will have a `r mean(tally)*100`% chance of not losing consecutive games in the 82-game season. But how precise is the estimate?

```{r variance}
var_phi = n_games * p_win*(1 - p_win)
mc_var = var_phi / S
```

The variance of the MC simulation is `r print(mc_var)`, which means the true estimate is the given percentage range with 95% probability: (`r (mean(tally) - sqrt(mc_var) *2) * 100`%, `r (mean(tally) + sqrt(mc_var) * 2) * 100`%).